Imposed form in the Early Acheulean? Evidence from Gona, Afar, Ethiopia

### 2023-12-01 manuscript/manuscript.Rmd
text update and figure reproduction.

Li, et al. [-@li2015] proposed a Flaked Area Index (FAI) as an alternative to SDI as a measure of reduction intensity, arguing that its validity is supported by an observed correlation (r=0.424) with SDI.
However, they also explain that "flaked area does not necessarily relate to the number of flake scars\...a small number of large scars can produce a large area of scar coverage, and conversely, a large number of small scars can produce a small area of scar coverage." (p. 6).
We suggest that what FAI actually captures is the spatial extent modifications to the surface of a core.
It is thus complementary to the measure of volume reduction provided by SDI and provides additional information to inform technological interpretations.
For example, a correlation between FAI and artifact form without any effect of SDI would suggest a focus on "least-effort" shape imposition whereas the opposite pattern would be consistent with relatively intense resharpening of spatially restricted areas on the core.
A lack of shape correlation with either measure would be expected for simple debitage with no morphological targets whereas a strong correlation with both would indicate a highly "designed" form achieved through extensive morphological and volumetric transformation.
In the current study we thus considered SDI and FAI together in order to evaluate evidence of intentional shaping in the early Acheulean of Gona.

@@

Conventional linear measures capture the direction (e.g., length \> breadth) but not the location of geometric relations (e.g., position of maximum breadth).
We address this by collecting linear measures defined by homologous semi-landmarks.
All artifacts were oriented along their maximum dimension, which was measured and defined as "length." The next largest dimension orthogonal to length was used to define the plane of "breadth," with the dimension orthogonal to this plane defined as "thickness" Width (W~1~, W~2~, W~3~) and thickness (T~1~, T~2~, T~3~) measures were then collected at 25%, 50%, and 75% of length, oriented so that 25% Width \> 75% Width.
To partition variation in shape from variation in size, we divided all linear measures by the geometric mean [@lycett2006].
GM-transformed variables were then submitted to a Principal Components Analysis (covariance matrix) to identify the main dimensions of shape variation.

@@

@@

Artifact scans (N=33) were cleaned, smoothed, and re-meshed using MeshLab.
The 3D triangular mesh of each scan was computed using Kazhdan and Hoppe's [-@kazhdan2013] method of screened Poisson surface reconstruction in MeshLab.
3DGM analysis was conducted using the *AGMT3-D* program of Herzlinger and Grosman [-@herzlinger2018].
Artifacts were automatically oriented according to the axis of least asymmetry, then manually oriented in the interests of standardization with the length, width, and thickness dimensions defined by the longest axis, followed by the next two longest axes perpendicular to the first.
The wider end of the artifacts was positioned as the proximal end (base), and more protruding surfaces were oriented towards the user in the first orthogonal view.
Then, a grid of 200 homologous semi-landmarks were overlain on each artifact's surface.
Generalized Procrustes and Principal Component analyses were then undertaken to explore the shape variability of the sample.
The surface area of each artifact was calculated using the *Artifact3-D* program of Grosman et al. [-@grosman2022].
*Artifact3-D* was also used to automatically identify the flake scar boundaries and compute each scar's surface area, using the scar analysis functions of Richardson et al. [-@richardson2014].

### Analyses

Measurement Validation: To assess the adequacy of shape descriptions based on our linear measures, we directly compared these with shape as quantified by 3D methods on the 33 artifacts for which scans are available.
GM-transformed linear measures from these 33 artifacts were submitted to a variance-covariance matrix PCA.
PCs with an eigenvalue greater than the mean were retained for analysis and the results compared qualitatively (morphological interpretation of PCs) and quantitatively (correlation of artifact factor scores) to 3D results.
Accuracy of surface area and flaked area estimates was also assessed by correlation with 3D results.

@@

A PCA on GM transformed linear measures the 33 artifacts identified two PCs accounting for 75.6% of the variance.
Linear-PC1 (Eigenvalue = 0.189) explained 55.2% of the variance.
Factor loadings (**Table** \@ref(tab:tab1)) for Linear-PC1 reflect artifact elongation (i.e., an anti-correlation of length vs. distal width and thickness).
This A PCA on GM transformed linear measures the 33 artifacts identified two PCs accounting for 75.6% of the variance.
Linear-PC1 (Eigenvalue = 0.189) explained 55.2% of the variance.
Factor loadings (**Table** \@ref(tab:tab1)) for Linear-PC1 reflect artifact elongation (i.e., an anti-correlation of length vs. distal width and thickness).
This closely parallels the length vs. width and thickness tradeoff captured by 3DGM-PC1 (Figure 3a) and is reflected in a tight correlation of artifact scores produced by the two PCs (r = 0.903, p \< 0.001, Figure 3c).
Comparison with conventional shape ratios (Table 2) similarly indicates that both Linear-PC1 and 3DGM-PC1 largely capture variation in artifact elongation.
A second factor (Linear-PC2, eigenvalue = 0.07) explained an additional 20.4% of variance.
This factor was less strongly correlated with its 3DGM counterpart (3DGM-PC2; r = 0.344, p = 0.050) probably because Linear-PC2 describes anticorrelated variation in width and thickness (i.e., broad and flat vs. thick and pointed; Table 1) whereas 3DGM-PC2 more purely isolates convergence (Figure 3a).
The remainder of the shape variability explained by Linear-PC2 is captured by higher order 3DGM-PCs like 3DGM-PC3 to 5, which comprise the contribution of the left and right lateral margins to relative thickness.
Use of high-resolution, coordinate-based scan data thus generates PCs that identify more specific shape attributes, but the underlying morphological variability captured by the linear and 3D analyses remains similar.
Together, 3DGM-PC2 (r = 0.344, p = 0.050), 3DGM-PC3 (r = -0.416, p = 0.016), 3DGM-PC4 (r = 0.458, p = 0.007), and 3DGM-PC5 (r = -0.352, p = 0.044) correlate well with Linear-PC2 (Figure 3c), cumulatively capturing whether the items are broad and flat or thick and pointed.
A stepwise regression (r2=0.625, F(4,28)=11.697, p\<0.001, Probability-of-F-to-enter \<= 0.050; Probability-of-F-to-remove \>= 0.100) with Linear-PC2 as the dependent variable retained all four of these 3DGM-PCs as significant predictors.

```{r tab1, echo=FALSE, message=FALSE}
library(tidyverse)
df1 <- data.frame("Linear metrics (GM transformed)" = c("Length", "W1", "W2", "W3", "T1", "T2", "T3"), 
                 "Linear-PC1" = c("0.989", "0.303", "0.403", "-0.176", "-0.135", "-0.369", "-0.607"), 
                 "Linear-PC2" = c("-0.107", "0.350", "0.767", "0.790", "-0.679", "-0.623", "-0.282") ) 
kableExtra::kbl(df1, booktabs = T, caption = "Component loadings for linear metric PCs .") 
```

We thus concluded that linear measures are adequate to capture relevant shape variation and proceeded with a PCA on our full sample.
We identified two PCs accounting for 80.0% of the variance.
PC1 (Eigenvalue = 0.216) explained 56.4% of the variance.
Factor loadings (Table 2) for PC1 reflect artifact flatness (i.e., an anti-correlation of length and width vs. thickness) such that higher values indicate relatively thinner pieces.
PC2 (Eigenvalue = 0.090) explained 23.6% of the variance.
Factor loadings (Table 2) show that PC2 captures artifact convergence or "pointedness" (i.e., an anti-correlation of tip width with length and butt thickness) such that higher values indicate shorter, less pointed forms.

We also tested the validity of our two reduction measures, SDI and FAI.
In agreement with Clarkson (2013), we found that surface area estimated from caliper measures displayed a strong correlation with (r2=0.975, p \< 0.001) but linear over-estimation of ($\beta$ = 1.58) 3DGM surface area (Figure 4a).
This results in a systematic underestimation of SDI that scales with core size (Figure 4b).
However, a simple correction of the caliper estimate (dividing by the slope, 1.58) eliminates surface area over-estimation (mean difference = 256mm2 [\<1.7% of mean], p=0.040) and produces SDI values that agree with 3DGM (r2=0.975, p \< 0.001, $\beta$ = 0.98) (Figure 4c).
We thus proceeded to apply this correction to surface area estimates in the full sample.
Insofar as these relationships are driven by basic geometry, we expect these methods (including correction) to be generalizable to other ESA assemblages.

@@

We first conducted a stepwise DFA on all flaked pieces (n=192) with inferred technological Mode (one vs. two) as the grouping variable and PCs 1 and 2, corrected SDI (cSDI), and FAI as the independent variables.
All variables were retained, yielding one canonical DF (eigenvalue=1.825, Wilks Lambda = 0.354, p\< 0.001) which correctly classified 93.8% of artifacts.
We thus accepted the validity of classification by Mode in our sample and employed this distinction in subsequent analyses.
There was no discernable difference in discriminant scores for Mode 1 cores from Oldowan (n=37) vs. Acheulean (n=39) contexts (, p = 0.746).
Mode 1 cores from Oldowan contexts (n=37) do include 10 (27%) small, retouched flakes.
Only one retouched flake from an Acheulean context was classified by the DFA as a Mode 1 core.
This piece, typologically classified as a "knife", is the smallest (93mm) retouched flake in the Acheulean sample.
When retouched flakes are excluded from the comparison, there are no significant differences in shape between Mode 1 cores from Oldowan and Achuelean contexts.
Interestingly, however, Acheulean Mode 1 cobble-cores are much larger (mean weight 480g vs. 186g, Cohen's d=1.137, p\<0.001) and less heavily reduced (mean cSDI 0.057 vs. 0.103, Cohen's d = -0.884, p \< 0.001) despite having similar FAI (mean 0.52 vs. 0.46, Cohen's d=0.271, p=0.266).

@@

In Mode 2 cores (n=115), reduction measures have different effects depending on initial blank form.
In general, the contrast between flat, lightly-worked flake bases and rounder, more-heavily worked cobble bases tends to inflate relationships between FAI and core shape (Figure 7c,d).
Mode 2 cobble cores (n=37) display a negative effect of FAI on PC1(Flatness) ($\beta$=-0.011, r=0.367, p = 0.026), no effect of FAI on pointedness, and no association between FAI and cSDI (r=0.141, p = 0.406).
Flake cores (n=69) show negative effects of FAI on PC1(Flatness) ($\beta$=-0.007, r=0.284, p = 0.018) and, more strongly, PC2(pointedness) ($\beta$=-0.019, r=0.443, p \< 0.001).
Flake cores show a small but significant positive effect of FAI on cSDI that, in contrast to Mode 1 cores, represents actual covariance rather than simply a relaxation of constraint (i.e. both upper and lower limits of cSDI increase with FAI).

Comparison of Mode 2 flake cores with unmodified large flakes from Achuelean contexts (n=35) shows that ranges of shape variation substantially overlap but that flaking generally reduces both PC1(flatness) and PC2(pointedness).
In fact, regressions of PC1 and PC2 on FAI show y-intercepts that closely approximate unmodified flake mean values (0.852 vs. 0.774 and 0.693 vs. 0.594, respectively) and significantly negative slopes as FAI increases (Figure 9).

A small number of Mode 2 cores (n=9) were so heavily modified (mean FAI=89%, range 65-100) that the blank form could not be determined.
These heavily worked "indeterminate" cores tend to show similar PC1(flatness) and PC2(pointedness) values to cobble-base cores (i.e. thicker and more pointed than flake cores; Figure x).
They also appear to follow general trends of shape change with increasing FAI observed across base types (Figure 10).

>>>>

Li, et al. [-@li2015] proposed a Flaked Area Index (FAI) as an alternative to SDI as a measure of reduction intensity, arguing that its validity is supported by an observed correlation (r=0.424) with SDI.
However, they also explain that "flaked area does not necessarily relate to the number of flake scars\...a small number of large scars can produce a large area of scar coverage, and conversely, a large number of small scars can produce a small area of scar coverage." (p. 6).
We suggest that what FAI actually captures is the spatial extent modifications to the surface of a core.
It is thus complementary to the measure of volume reduction provided by SDI and provides additional information to inform technological interpretations.
For example, a correlation between FAI and artifact form without any effect of SDI would suggest a focus on "least-effort" shape imposition whereas the opposite pattern would be consistent with relatively intense resharpening of spatially restricted areas on the core.
A lack of shape correlation with either measure would be expected for simple debitage with no morphological targets whereas a strong correlation with both would indicate a highly "designed" form achieved through extensive morphological and volumetric transformation. (**Figure** \@ref(fig:FAISDI))
In the current study we thus considered SDI and FAI together in order to evaluate evidence of intentional shaping in the early Acheulean of Gona.

```{r FAISDI, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="A conceptual model of inferring intention from artifact morphology."}
knitr::include_graphics(here::here("figure/FAI_SDI.png"))
```

@@

Conventional linear measures capture the direction (e.g., length \> breadth) but not the location of geometric relations (e.g., position of maximum breadth).
We address this by collecting linear measures defined by homologous semi-landmarks.
All artifacts were oriented along their maximum dimension, which was measured and defined as "length". The next largest dimension orthogonal to length was used to define the plane of "breadth", with the dimension orthogonal to this plane defined as "thickness" Width (W~1~, W~2~, W~3~) and thickness (T~1~, T~2~, T~3~) measures were then collected at 25%, 50%, and 75% of length, oriented so that 25% Width \> 75% Width.
To partition variation in shape from variation in size, we divided all linear measures by the geometric mean [@lycett2006].
GM-transformed variables were then submitted to a Principal Components Analysis (covariance matrix) to identify the main dimensions of shape variation.

@@

To assess the adequacy of shape descriptions based on our linear measures, we directly compared these with shape as quantified by 3D methods on the 33 artifacts for which scans are available. GM-transformed linear measures from these 33 artifacts were submitted to a variance-covariance matrix PCA. PCs with an eigenvalue greater than the mean were retained for analysis and the results compared qualitatively (morphological interpretation of PCs) and quantitatively (correlation of artifact factor scores) to 3D results. Accuracy of surface area and flaked area estimates was also assessed by correlation with 3D results.

@@

3DGM analysis was conducted using the *AGMT3-D* program of Herzlinger and Grosman [-@herzlinger2018].
Artifacts were automatically oriented according to the axis of least asymmetry, then manually oriented following the same orthogonal conventions described in section 2.3. Then, a grid of 200 homologous semi-landmarks were overlain on each artifact's surface. Generalized Procrustes and Principal Component analyses were then undertaken to explore the shape variability of the sample. The surface area of each artifact was calculated using the *Artifact3-D* program of Grosman et al. [-@grosman2022]. *Artifact3-D* was also used to automatically identify the flake scar boundaries and compute each scar's surface area, using the scar analysis functions of Richardson et al. [-@richardson2014].

@@

A PCA on GM transformed linear measures of the 33 artifacts identified two PCs accounting for 75.6% of the variance.
Linear-PC1 (Eigenvalue = 0.189) explained 55.2% of the variance.
Factor loadings (**Table** \@ref(tab:tab1)) for Linear-PC1 reflect artifact elongation (i.e., an anti-correlation of length vs. distal width and thickness).
This A PCA on GM transformed linear measures the 33 artifacts identified two PCs accounting for 75.6% of the variance.
Linear-PC1 (Eigenvalue = 0.189) explained 55.2% of the variance.
Factor loadings (**Table** \@ref(tab:tab1)) for Linear-PC1 reflect artifact elongation (i.e., an anti-correlation of length vs. distal width and thickness).
This closely parallels the length vs. width and thickness tradeoff captured by 3DGM-PC1 (**Figure** \@ref(fig:fig3d)) and is reflected in a tight correlation of artifact scores produced by the two PCs (r = 0.903, p \< 0.001, **Figure** \@ref(fig:figcomparison)A).
A second factor (Linear-PC2, eigenvalue = 0.07) explained an additional 20.4% of variance. This factor was less strongly correlated with its 3DGM counterpart (3DGM-PC2; r = 0.344, p = 0.050) probably because Linear-PC2 describes anticorrelated variation in width and thickness (i.e., broad and flat vs. thick and pointed; **Table** \@ref(tab:tab1)) whereas 3DGM-PC2 more purely isolates convergence (**Figure** \@ref(fig:fig3d)). The remainder of the shape variability explained by Linear-PC2 is captured by higher order 3DGM-PCs 3 through 5, which comprise the contribution of the left and right lateral margins to relative thickness. Use of high-resolution, coordinate-based scan data thus generates PCs that identify more specific shape attributes, but the underlying morphological variability captured by the linear and 3D analyses remains similar. Together, 3DGM-PC2 (r = 0.344, p = 0.050), 3DGM-PC3 (r = -0.416, p = 0.016), 3DGM-PC4 (r = 0.458, p = 0.007), and 3DGM-PC5 (r = -0.352, p = 0.044) correlate well with Linear-PC2, cumulatively capturing whether the items are broad and flat or thick and pointed. A stepwise regression ($r^2$=0.625, F(4,28)=11.697, p<0.001, Probability-of-F-to-enter <= 0.050; Probability-of-F-to-remove >= 0.100) with Linear-PC2 as the dependent variable retained all four of these 3DGM-PCs as significant predictors.

```{r tab1, echo=FALSE, message=FALSE}
library(tidyverse)
df1 <- data.frame("Linear metrics (GM transformed)" = c("Length", "W1", "W2", "W3", "T1", "T2", "T3"), 
                 "Linear-PC1" = c("0.989", "0.303", "0.403", "-0.176", "-0.135", "-0.369", "-0.607"), 
                 "Linear-PC2" = c("-0.107", "0.350", "0.767", "0.790", "-0.679", "-0.623", "-0.282") ) 
kableExtra::kbl(df1, booktabs = T, caption = "Component loadings for linear metric PCs on scanned sample.") 
```

```{r fig3d, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="3D models displaying the first two PCs."}
knitr::include_graphics(here::here("figure/Fig3a.png"))
```

```{r figcomparison, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="Comparison between linear measurement and 3d measurement."}
knitr::include_graphics(here::here("figure/Fig.3D comparison.png"))
```

We thus concluded that linear measures are adequate to capture relevant shape variation and proceeded with a PCA on our full sample.
We identified two PCs accounting for 80.0% of the variance.
PC1 (Eigenvalue = 0.216) explained 56.4% of the variance.
Factor loadings (**Table** \@ref(tab:tab2)) for PC1 reflect artifact flatness (i.e., an anti-correlation of length and width vs. thickness) such that higher values indicate relatively thinner pieces.
PC2 (Eigenvalue = 0.090) explained 23.6% of the variance.
Factor loadings (**Table** \@ref(tab:tab2)) show that PC2 captures artifact convergence or "pointedness" (i.e., an anti-correlation of tip width with length and butt thickness) such that higher values indicate shorter, less pointed forms.

```{r tab2, echo=FALSE, message=FALSE}
library(tidyverse)
df1 <- data.frame("Linear metrics (GM transformed)" = c("Length", "W1", "W2", "W3", "T1", "T2", "T3"), 
                 "PC1" = c("0.905", "0.635", "0.680", "0.486", "-0.510", "-0.684", "-0.719"), 
                 "PC2" = c("-0.414", "0.299", "0.571", "0.761", "-0.526", "-0.525", "0.023") ) 
kableExtra::kbl(df1, booktabs = T, caption = "Component loadings for linear metric PCs in the full sample.") 
```

We also tested the validity of our two reduction measures, SDI and FAI.
In agreement with Clarkson (2013), we found that surface area estimated from caliper measures displayed a strong correlation with (r2=0.975, p \< 0.001) but linear over-estimation of ($\beta$ = 1.58) 3D surface area.
This results in a systematic underestimation of SDI that scales with core size. However, a simple correction of the caliper estimate (dividing by the slope, 1.58) eliminates surface area over-estimation (mean difference = $256mm^2$ [\<1.7% of mean], p=0.040) and produces SDI values that agree with 3D values (r2=0.975, p \< 0.001, $\beta$ = 0.98) (**Figure** \@ref(fig:figcomparison)B).
We thus proceeded to apply this correction to surface area estimates in the full sample.
Insofar as these relationships are driven by basic geometry, we expect these methods (including correction) to be generalizable to other ESA assemblages.

Visual estimation of flaked area approximated 3DGM measurement very closely (r2=0.932, p < 0.001, $\beta$ = 1.051; **Figure** \@ref(fig:figcomparison)C) and without any systematic bias (paired t-test: mean difference = -0.015%, 95% CI = 2.02% to 1.99%, p = 0.987). Individual errors ranged between -10.46% and 14.24%. We thus considered visual estimation to be reliable in our sample.

@@

We first conducted a stepwise DFA on all flaked pieces (n=192) with inferred technological Mode (one vs. two) as the grouping variable and PCs 1 and 2, corrected SDI (cSDI), and FAI as the independent variables.
All variables were retained, yielding one canonical DF (eigenvalue=1.825, Wilks Lambda = 0.354, p\< 0.001) which correctly classified 93.8% of artifacts.
We thus accepted the validity of classification by Mode in our sample and employed this distinction in subsequent analyses.
There was no discernable difference in discriminant scores for Mode 1 cores from Oldowan (n=37) vs. Acheulean (n=39) contexts (, p = 0.746).
Mode 1 cores from Oldowan contexts (n=37) do include 10 (27%) small, retouched flakes.
Only one retouched flake from an Acheulean context was classified by the DFA as a Mode 1 core.
This piece, typologically classified as a "knife", is the smallest (93mm) retouched flake in the Acheulean sample.
When retouched flakes are excluded from the comparison, there are no significant differences in shape between Mode 1 cores from Oldowan and Acheulean contexts.
Interestingly, however, Acheulean Mode 1 cobble-cores are much larger (mean weight 480g vs. 186g, Cohen's d=1.137, p\<0.001) and less heavily reduced (mean cSDI 0.057 vs. 0.103, Cohen's d = -0.884, p \< 0.001) despite having similar FAI (mean 0.52 vs. 0.46, Cohen's d=0.271, p=0.266).

@@

In Mode 2 cores (n=115), reduction measures have different effects depending on initial blank form.
In general, the contrast between flat, lightly-worked flake bases and rounder, more-heavily worked cobble bases tends to inflate relationships between FAI and core shape (**Figure** \@ref(fig:FAIBASE)).
Mode 2 cobble cores (n=37) display a negative effect of FAI on PC1(Flatness) ($\beta$=-0.011, r=0.367, p = 0.026), no effect of FAI on pointedness, and no association between FAI and cSDI (r=0.141, p = 0.406).
Flake cores (n=69) show negative effects of FAI on PC1(Flatness) ($\beta$=-0.007, r=0.284, p = 0.018) and, more strongly, PC2(pointedness) ($\beta$=-0.019, r=0.443, p \< 0.001). Comparison of Mode 2 flake cores with unmodified large flakes from Acheulean contexts (n=35) shows that ranges of shape variation substantially overlap but that flaking generally reduces both PC1 and PC2. In fact, regressions of PC1 and PC2 on FAI show y-intercepts that closely approximate unmodified flake mean values (0.852 vs. 0.774 and 0.693 vs. 0.594, respectively) and significantly negative slopes as FAI increases (**Figure** \@ref(fig:FAIFLAKED)). 
Flake cores show a small but significant positive effect of FAI on cSDI that, in contrast to Mode 1 cores, represents actual covariance rather than simply a relaxation of constraint (i.e. both upper and lower limits of cSDI increase with FAI).

```{r FAIBASE, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="Mode 2 trends across blank types."}
knitr::include_graphics(here::here("figure/Fig.FAI by base.png"))
```

```{r FAIFLAKED, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="Effect of FAI compared to unmodified large flakes from Acheulean contexts."}
knitr::include_graphics(here::here("figure/Fig.FAI by flaked.png"))
```

A small number of Mode 2 cores (n=9) were so heavily modified (mean FAI=89%, range 65-100) that the blank form could not be determined.
These heavily worked "indeterminate" cores tend to show similar PC1 (flatness) and PC2 (pointedness) values to cobble-base cores (i.e. thicker and more pointed than flake cores) and follow general trends of shape change with increasing FAI observed across base types (**Figure** \@ref(fig:FAIBASE)).
